ISYE 6740 HW 1

### 2024-08-29 msiddiqui60_HW1_report.tex
Added Problem 4

>>>>

Third image source: https://commons.wikimedia.org/wiki/File:Greycat.jpg

The code for the Image Compression is in the Jupyter Notebook, as are the output images and original images.
I originally wrote a k-means algorithm which would randomly re-assign centroid location for empty clusters and this
performed well with respect to visual appearance of the rendered images. To bring my submission more in line
with the requirements for the assignment, I implemented a change such that during a given iteration, any empty
cluster was dropped. For low number k, this was't an issue, however, for k = 16 and k = 32, all three images
exhibit significant drops, with final number of clusters anywhere from 11 to 25 respectively. I could visually
tell that the image was not as crisp as the original nor the rendered images with re-assigned clusters. Additionally,
the time it took to run my function in both cases was roughly the same about 2 minutes to 3 minutes for the
football image depending on the set of random numbers I used. Later, I added a seed of 123 for reproducibility.

Additionally, although I originally implemented a loop to calculate the distance and assign points to centroids,
I then used numpy calculations before switching entirely to using the lin.alg.norm built-in function to help
calculate the l2_norm squared or euclidian squared distance. I've commented out the originals to show the implementation.

For each k, I ran 5 runs and selected the best run by minimizing the within cluster sum of squares metric.
Per Datanovia (source link below), the within cluster sum of squares is a method capturing within cluster variance
and captures the "compactness" of clusters. To calculate, the squared difference of the distance between
data points and centroids are calculated and then summed for all clusters. A lower WCSS indicates that points
tend to be closer to their centroids and are therefore better fit. In our runs, the iterations which on average
had the lowest WCSS or the best fitting clusters were selected for each k. The kmeans package in R allows you
to set the number of initial starts parameter nstart which is highly recommended. You can do 25, 50 or many more
runs. In doing so, you may develop more stable results by selecting the variant with the lowest WCSS.

The higher the number of centroids k, the longer the algorithm took to run. For the football image, the best
run for k = 2 took 0.36 seconds, while the best run for k = 32 took 15.15 total seconds. This makes
sense as increasing the number of k makes running the algorithm computationally expensive. Similarly, the number
of iterations kmeans went through before converging also increased as the k did. For the football image,
when k = 2, I only need 7 iterations for the best run, whereas when k = 32, I required 42 iterations for the best
run. This makes sense as having more clusters would take longer to converge.

I returned multiple data points in my kmeans function which I printed out as part of the run_k_means_and_visualize
function. These included the length of the best iteration, k, number of centroids, best initialized centroids,
and the number of iterations for each run as well as the best image itself. You can see the rendered images
for each specified k as well as the three original images below.

As for finding the best k for this specific image compression issue, this depends entirely on your goals. The lower
the number of clusters k, the more compressed the image will be and the less time it will take to run the algorithm.
However, in doing so, the visual representation will not be as crisp or clear as a higher k rendered image. It is
important to determine what the situation calls for in determining how much to compress the image. In this instance,
k = 32 provided the best rendered image though it may take up more space than the lower k images. Additionally,
at 15 seconds a run, it didn't take too long. However, in other situations, determining the optimal number of clusters
is wholly determined by the problem. One method used to determine optimal k is the elbow method. where one may
plot the WCSS for different values of k and then find the "elbow" or sharp flattening of the curve which will
correspond to an optimal k. Another method is to simply use domain knowledge or visually plot the data (when possible)
to highlight existing clusters. There are other methods such as Silhouette analysis etc. For image compression,
a visual cue along with examing tradeoff between visual acuity, size, and time to compute + render are the
tradeoffs we would examine as the image will become more crisp as k increases.

\subsection{Football}
\begin{figure}
    \centering
    \includegraphics{c/users/moham/pycharmprojects/hw1_isye_6740/data/football}
    \caption{Original}
    \label{fig:football}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{C:\Users\moham\PycharmProjects\HW1_ISYE_6740\outputs\football_2}
    \caption{K = 2}
    \label{fig:football_2}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{C:\Users\moham\PycharmProjects\HW1_ISYE_6740\outputs\football_4}
    \caption{K = 4}
    \label{fig:football_4}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{C:\Users\moham\PycharmProjects\HW1_ISYE_6740\outputs\football_8}
    \caption{K = 8}
    \label{fig:football_8}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{C:\Users\moham\PycharmProjects\HW1_ISYE_6740\outputs\football_16}
    \caption{K = 16}
    \label{fig:football_16}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{C:\Users\moham\PycharmProjects\HW1_ISYE_6740\outputs\football_32}
    \caption{K = 32}
    \label{fig:football_32}
\end{figure}

\subsection{Cat}
\begin{figure}
    \centering
    \includegraphics{C:\Users\moham\PycharmProjects\HW1_ISYE_6740\data\256px-Greycat}
    \caption{Original}
    \label{fig:256px-greycat}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{C:\Users\moham\PycharmProjects\HW1_ISYE_6740\outputs\256px-Greycat_2}
    \caption{K = 2}
    \label{fig:256px-greycat_2}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{C:\Users\moham\PycharmProjects\HW1_ISYE_6740\outputs\256px-Greycat_4}
    \caption{K = 4}
    \label{fig:256px-greycat_4}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{C:\Users\moham\PycharmProjects\HW1_ISYE_6740\outputs\256px-Greycat_8}
    \caption{K = 8}
    \label{fig:256px-greycat_8}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{C:\Users\moham\PycharmProjects\HW1_ISYE_6740\outputs\256px-Greycat_16}
    \caption{K = 16}
    \label{fig:256px-greycat_16}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{C:\Users\moham\PycharmProjects\HW1_ISYE_6740\outputs\256px-Greycat_32}
    \caption{K = 32}
    \label{fig:256px-greycat_32}
\end{figure}

\subsection{Coastal Abstract}

\begin{figure}
    \centering
    \includegraphics{C:\Users\moham\PycharmProjects\HW1_ISYE_6740\data\coastal-abstract}
    \caption{Original}
    \label{fig:coastal-abstract}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{C:\Users\moham\PycharmProjects\HW1_ISYE_6740\outputs\coastal-abstract_2}
    \caption{K = 2}
    \label{fig:coastal-abstract_2}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{C:\Users\moham\PycharmProjects\HW1_ISYE_6740\outputs\coastal-abstract_4}
    \caption{K = 4}
    \label{fig:coastal-abstract_4}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{C:\Users\moham\PycharmProjects\HW1_ISYE_6740\outputs\coastal-abstract_8}
    \caption{K = 8}
    \label{fig:coastal-abstract_8}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{C:\Users\moham\PycharmProjects\HW1_ISYE_6740\outputs\coastal-abstract_16}
    \caption{K = 16}
    \label{fig:coastal-abstract_16}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{C:\Users\moham\PycharmProjects\HW1_ISYE_6740\outputs\coastal-abstract_32}
    \caption{K = 32}
    \label{fig:coastal-abstract_32}
\end{figure}

\subsection{Sources}

\begin{enumerate}
\item https://www.datanovia.com/en/lessons/k-means-clustering-in-r-algorith-and-practical-examples/
\item https://www.geeksforgeeks.org/find-a-matrix-or-vector-norm-using-numpy/#
\item https://www.w3schools.com/python/numpy/numpy_creating_arrays.asp
\item https://numpy.org/doc/stable/reference/generated/numpy.zeros.html
\item https://numpy.org/devdocs/reference/constants.html#numpy.newaxis
\item https://note.nkmk.me/en/python-opencv-bgr-rgb-cvtcolor/
\item https://stackoverflow.com/questions/49434754/how-does-the-pyplot-imshow-function-work
\item https://www.geeksforgeeks.org/ml-determine-the-optimal-value-of-k-in-k-means-clustering/

\end{enumerate}

\subsection{Steps Taken for Algorithm}
Steps Taken:
- convert images into pixel color representations
- define the number of clusters
- set the initial position of each cluster randomly (you are going to run this multiple times)
- use the l2-norm squared distance formula for each pixel to assign a cluster
- update the clusters
- repeat assignment until convergence

\section{MNIST Dataset Clustering}

Steps I took:
- load in the MNIST file
- only maintain the train data set (per instructions)
- preprocess per instructions
- write (or import) the k-means with eucldiean squared distance
- write (or import) the k-means with manhattan squared distance
- calculate clusters for each image/data point
- assign class (from labeled y data) to each cluster based on which class is most common for a given cluster
- calculate purity score for both kmeans runs for each cluster as well as average purity score

\subsection{Results and Discussion}
Please reference the added Jupyter notebook for code and direct output. I used the pyclustering library
implementation of kmeans so that I could easily use both Manhattan and squared l2norm distance metrics.
Additionally, I implemented the kmeans++ methodolgy of setting initial centroids to avoid needing to run each
algorithm multiple times and choose the run with the lowest WCSS as I did for section 3. Following the steps
I took outlined in the above (standardizing, dropping test sets, assigning clusters, finding the majority
labels, anc calculating purity score as shown below), I was able to obtain average purity score of roughly 58.28
percent when using squared euclidean as the distance metric and 51.69 when using Manhattan distance as the distance
metric. This suggests that the squared euclidean distance metric captures the MNIST data structure more effectively
as it is more sensitive to large differences in distances between data points than manhattan distance.
\[
\mbox{purity}_i = \frac{\mbox{corrected assigned samples}_i}{\mbox{size of cluster}_i}
\]
Below are the individual purity scores and assigned label for each cluster for the Euclidean Squared distance
run of the K-Means algorithm

\begin{tabular}{|c|c|c|}
\hline
\textbf{Cluster} & \textbf{Assigned Label} & \textbf{Purity Score} \\
\hline
0 & 1 & 0.5902 \\
\hline
1 & 0 & 0.9385 \\
\hline
2 & 3 & 0.5310 \\
\hline
3 & 5 & 0.3058 \\
\hline
4 & 6 & 0.8723 \\
\hline
5 & 1 & 0.6471 \\
\hline
6 & 4 & 0.3585 \\
\hline
7 & 8 & 0.6194 \\
\hline
8 & 2 & 0.9177 \\
\hline
9 & 7 & 0.4367 \\
\hline
\end{tabular}

Below are the individual purity scores and assigned label for each cluster for the Manhattan distance
run of the K-Means algorithm.

\begin{table}[h!]
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Cluster} & \textbf{Assigned Label} & \textbf{Purity Score} \\
\hline
0 & 8 & 0.5693 \\
1 & 8 & 0.4366 \\
2 & 1 & 0.3586 \\
3 & 7 & 0.3841 \\
4 & 6 & 0.8728 \\
5 & 0 & 0.9410 \\
6 & 3 & 0.5966 \\
7 & 2 & 0.9603 \\
8 & 4 & 0.3948 \\
9 & 0 & 0.8901 \\
\hline
\end{tabular}
\caption{Cluster Assignments and Purity Scores}
\label{tab:cluster_purity}
\end{table}

\end{document}
